hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-02-14 inst/manuscript/paper.tex

% --- Header / Footer ---------------------------------------------------------
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
%\renewcommand{\headrulewidth}{0.50pt}
\renewcommand{\headrulewidth}{0pt}
\fancyhead[L]{\hspace{-0.75cm}\includegraphics[width=5.5cm]{C:/Users/BryanMaitland/AppData/Local/R/win-library/4.4/rticles/rmarkdown/templates/joss/resources/JOSS-logo.png}}
\fancyhead[C]{}
\fancyhead[R]{}
\renewcommand{\footrulewidth}{0.25pt}

@@

@@

Understanding the timing of key life history events is necessary for
managing and conserving populations. Historically, models to predict
hatch and emergence timing for fishes were difficult to employ in wild
settings because average incubation temperature was needed as the
primary parameter in predictive models. However, recent improvements to
these techniques reworked models such that they could be applied in wild
environments as long as users had data for when adult fish spawned and a
record of average daily temperature over the course of development.
Despite these improvements, their application remains limited due to few
parameterizations for different species, being largely limited to
salmonids. Here we present \texttt{hatchR}, a software ecosystem that
allows users to predict hatch and emergence timing for wild fishes, as
well as additional tools to aid in those analyses. \texttt{hatchR}
allows users to leverage popular historic parameterizations for
phenological models or to easily implement custom parameterizations
using data not included in the package. \texttt{hatchR} is also
distributed in two forms---an open source R package for maximum
customization, as well as an HTML graphical-user-interface web
application for individuals not familiar with scripting languages. To
demonstrate potential uses, we present two case studies as likely
applications for this software. \texttt{hatchR} promises to open many
exciting avenues in research and management of fishes during their early
life history.

@@

As primarily poikilothermic organisms, the development and growth of
fishes is tightly linked with the temperature of their ambient
environment. This close relationship has allowed researchers to generate
statistical models that allow the prediction of developmental phenology
with high accuracy and precision. These models were typically developed
in aquaculture settings and their initial formulations were not
applicable to wild populations because they assumed a constant
temperature over the course of development Beacham \& Murray (1990)
(\textbf{add more!}) . However, Sparks et al. (2019) reformulated this
approach as an ``Effective Value model'', in which the input was daily
average temperature after a parent spawned and fish would either hatch
or emerge when effective values cumulatively summed to one.

The resulting effective value approach has now been widely applied in
salmonids for which parameterizations from aquaculture were readily
available---for example Pacific Salmon (\emph{Oncorhynchus spp.}) models
developed by Beacham \& Murray (1990) were have been applied to various
species and populaions (Adelfio, Wondzell, Mantua, \& Reeves, 2019,
2024; Kaylor et al., 2021) while models developed for Bull Trout
(\emph{Salvelinus confluentus}) by McPhail \& Murray (1979) were
extended by Austin, Essington, \& Quinn (2019). Despite growing
popularity, applications have been largely limited within Salmonids,
presumably because parameterizations for such models already existed due
to their wide use in aquaculture and their general popularity as sport
and commercial fish.

To bridge the gap between the application of one-off effective value
model applications within individual studies and the lack of
parameterization for other species, we developed the software ecosystem,
\texttt{hatchR}. Specifically, \texttt{hatchR} allows users to input
standard raw or summarized temperature datasets that are commonly
collected in wild settings, run basic checks on those data, use built-in
parameterizations like those from Beacham \& Murray (1990) or Sparks,
Westley, Falke, \& Quinn (2017), develop custom models from their own
temperature and phenological data, and predict hatch and emergence
timing using these models in the effective value framework.

To widen the user application of these methods, we distribute two
user-interfaces for \texttt{hatchR}. The first is a R package
distributed via CRAN that allows users the most customizable application
of these methods. The R package is especially powerful as it allows
users to automate their analyses over multiple variables such as
phenology type, multiple spawn dates, or different habitats with varying
thermal regimes. These variable approaches are outlined in the package
documentation on \texttt{hatchR}'s website. Alternatively, we also
distribute a Shiny application in the form of an HTML-based web tool to
interact with many of \texttt{hatchR}'s functions in a
graphical-user-interface. The Shiny form trades-off some of automative
power for user simplicity, while still allowing users to leverage much
of the functionality of \texttt{hatchR}'s R package. Below, we present
the basic overview of the software and multiple case studies of how it
may be applied.

@@

\texttt{hatchR} is meant to primarily be a tool for predicting
phenology. In this sense, we mostly limit functionality to these
applications and provide minimal data checking and plotting help. This
decision is in part driven by the diversity of data types that users may
import and the difficulty in addressing all those data types with
respect to various data checks. In other words, we expect users to know
their data better than we do and to check it accordingly. We do provide
two basic data check functions discussed in the Checking Data section.
Similarly, we provide limited functionality for plotting results, but
provide examples of how to build custom visualization from output,
specifically in R. For the Shiny application, we provide a base output
plot, but the ability to download your results for custom plotting in
programs of the user's choice.

\includegraphics{flowchart.png}

@@

Effective value models were created by Sparks et al. (2019) to implement
developmental models in wild environments for Sockeye Salmon (\emph{O.
nerka}). The need for change arose because historic models, specifically
those in Beacham \& Murray (1990), only considered the average
incubation temperature during development and, for wild fishes, average
incubation temperature was impossible to estimate because it was unknown
when fish hatched even if adult spawn timing was known. To address this,
Sparks et al. (2019) used the reciprocal of the formulation of model 2
from Beacham \& Murray (1990) and assigned an effective value for every
day of development using the daily average temperature.

@@

Where \emph{i} is the daily value and a fish hatches or emerges when the
cumulative sum or \[\sum_{i =1}^nEffectiveValue_i = 1\]

The effective value model framework is the basis for the phenological
models in \texttt{hatchR}, both in the included \texttt{model\_table} in
the package, as well as for custom models users can fit with
\texttt{fit\_model()}. Specifically, \texttt{model\_table} has been
extended to include more parameterizations from Beacham \& Murray
(1990), Sparks et al. (2017), and Austin et al. (2019) (who extended
McPhail \& Murray (1979)).

@@

@@

\texttt{hatchR} is built assuming data will be analyzed as daily average
temperatures. Despite that assumption, raw data (\emph{e.g.}, as
outputted by HOBO loggers) can be used and \texttt{hatchR} includes
functionality to summarize those data into a format that is usable, as
well as provides functions for basic visual and programmatic data checks
to make sure outliers or missing data are at least brought to users'
attention.

summarize\_temp, check\_continuous and plot\_check\_temp

@@

talk about using nls for fit\_model

model\_table and fit\_model

fit\_model for three non-salmonid species

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{smallmouth }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{2}\NormalTok{) }\SpecialCharTok{|\textgreater{}} \FunctionTok{tibble}\NormalTok{()}
\FunctionTok{colnames}\NormalTok{(smallmouth) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"hours"}\NormalTok{, }\StringTok{"temp\_F"}\NormalTok{)}
\NormalTok{smallmouth}\SpecialCharTok{$}\NormalTok{hours }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{52}\NormalTok{, }\DecValTok{54}\NormalTok{, }\DecValTok{70}\NormalTok{, }\DecValTok{78}\NormalTok{, }\DecValTok{90}\NormalTok{, }\DecValTok{98}\NormalTok{, }\DecValTok{150}\NormalTok{, }\DecValTok{167}\NormalTok{, }\DecValTok{238}\NormalTok{, }\DecValTok{234}\NormalTok{)}
\NormalTok{smallmouth}\SpecialCharTok{$}\NormalTok{temp\_F }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{77}\NormalTok{, }\DecValTok{75}\NormalTok{, }\DecValTok{71}\NormalTok{, }\DecValTok{70}\NormalTok{, }\DecValTok{67}\NormalTok{, }\DecValTok{65}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{59}\NormalTok{, }\DecValTok{55}\NormalTok{, }\DecValTok{55}\NormalTok{)}

\NormalTok{smallmouth }\OtherTok{\textless{}{-}}\NormalTok{ smallmouth }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{days =} \FunctionTok{ceiling}\NormalTok{(hours}\SpecialCharTok{/}\DecValTok{24}\NormalTok{),}
         \AttributeTok{temp\_C =}\NormalTok{ (temp\_F }\SpecialCharTok{{-}}\DecValTok{32}\NormalTok{) }\SpecialCharTok{*}\NormalTok{ (}\DecValTok{5}\SpecialCharTok{/}\DecValTok{9}\NormalTok{))}

\FunctionTok{plot}\NormalTok{(smallmouth}\SpecialCharTok{$}\NormalTok{days}\SpecialCharTok{\textasciitilde{}}\NormalTok{smallmouth}\SpecialCharTok{$}\NormalTok{temp\_C)}

\NormalTok{smb\_mod }\OtherTok{\textless{}{-}} \FunctionTok{fit\_model}\NormalTok{(smallmouth,}
                     \AttributeTok{temp =}\NormalTok{ smallmouth}\SpecialCharTok{$}\NormalTok{temp\_C,}
                     \AttributeTok{days =}\NormalTok{ smallmouth}\SpecialCharTok{$}\NormalTok{hours)}
\end{Highlighting}
\end{Shaded}

@@

Use woody \_example from website

show predict\_phenology output slots

show plot\_phenology

>>>>

% --- Header / Footer ---------------------------------------------------------
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
%\renewcommand{\headrulewidth}{0.50pt}
\renewcommand{\headrulewidth}{0pt}
\fancyhead[L]{\hspace{-0.75cm}\includegraphics[width=5.5cm]{/Library/Frameworks/R.framework/Versions/4.4-arm64/Resources/library/rticles/rmarkdown/templates/joss/resources/JOSS-logo.png}}
\fancyhead[C]{}
\fancyhead[R]{}
\renewcommand{\footrulewidth}{0.25pt}

@@

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

@@

Understanding the timing of key life history events is necessary for
managing and conserving populations. Historically, models to predict
hatch and emergence timing for fishes were difficult to employ in wild
settings because average incubation temperature was needed as the
primary parameter in predictive models. However, recent improvements to
these techniques reworked models such that they could be applied in wild
environments as long as users had data for when adult fish spawned and a
record of average daily temperature over the course of development.
Despite these improvements, their application remains limited due to few
parameterizations for varying species, being largely limited to
salmonids. Here we present \textbf{hatchR}, a software ecosystem that
allows users to predict hatch and emergence timing for wild fishes, as
well as additional tools to aid in those analyses. \textbf{hatchR}
allows users to leverage popular historic parameterizations for
phenological models or to easily implement custom parameterizations
using data not included in the package. \textbf{hatchR} is also
distributed in two forms---an open source R package for maximum
customization, as well as an HTML graphical-user-interface web
application for individuals not familiar with scripting languages. To
demonstrate potential uses, we present two case studies as likely
applications for this software. \textbf{hatchR} promises to open many
exciting avenues in research and management of fishes during their early
life history.

@@

As primarily poikilothermic organisms, the development and growth of
fishes is tightly linked with the temperature of their ambient
environment. This close relationship has allowed researchers to generate
statistical models that allow the prediction of developmental phenology
with high accuracy and precision. These models were typically developed
in aquaculture settings and their initial formulations were not
applicable to wild populations because they assumed a constant
temperature over the course of development (Alderdice \& Velsen, 1978;
Beacham \& Murray, 1990; McPhail \& Murray, 1979). However, Sparks et
al. (2019) reformulated this approach as an ``Effective Value model'',
in which the input was daily average temperature after a parent spawned
and fish would either hatch or emerge when effective values cumulatively
summed to one.

The resulting effective value approach has now been widely applied in
Salmonids for which parameterizations from aquaculture were readily
available---for example Pacific Salmon (\emph{Oncorhynchus spp.}) models
developed by Beacham \& Murray (1990) have been applied to various
species and populations (Adelfio, Wondzell, Mantua, \& Reeves, 2019,
2024; Kaylor et al., 2021) while models developed for Bull Trout
(\emph{Salvelinus confluentus}) by McPhail \& Murray (1979) were
extended by Austin, Essington, \& Quinn (2019). Despite growing
popularity, applications have been largely limited within Salmonids,
presumably because parameterizations for such models already existed due
to their wide use in aquaculture and their general popularity as sport
and commercial fish.

To bridge the gap between the application of one-off effective value
model applications within individual studies and the lack of
parameterization for other species, we developed the software ecosystem,
\textbf{hatchR}. Specifically, \textbf{hatchR} allows users to input
standard raw or summarized temperature datasets that are commonly
collected in wild settings, run basic checks on those data, use built-in
parameterizations like those from Beacham \& Murray (1990) or Sparks,
Westley, Falke, \& Quinn (2017), develop custom models from their own or
published temperature and phenological data, and predict hatch and
emergence timing using these models in the effective value framework.

To widen the user application of these methods, we distribute two
user-interfaces for \textbf{hatchR}. The first is a R package ({``R''}
(n.d.)) distributed via CRAN that allows users the most customizable
application for these methods. The R package is especially powerful as
it allows users to automate their analyses over multiple variables such
as phenology type, multiple spawn dates, or different habitats with
varying thermal regimes. These variable approaches are outlined in the
package documentation on \textbf{hatchR}'s website. Alternatively, we
also distribute a Shiny application (Chang et al. (2024)) in the form of
an HTML-based web tool to interact with many of \textbf{hatchR}'s
functions in a graphical-user-interface. The Shiny form trades-off some
of automative power for user simplicity, while still allowing users to
leverage much of the functionality of \textbf{hatchR}'s R package.
Below, we present the basic overview of the software and multiple case
studies of how it may be applied.

@@

\textbf{hatchR} is meant to primarily be a tool for predicting
phenology. In this sense, we limit functionality to these applications
and provide minimal data checking and plotting help. This decision is in
part driven by the diversity of data types that users may import and the
difficulty in addressing all those data types with respect to various
data checks. In other words, we expect users to know their data better
than we do and to check it accordingly. We do provide two basic data
check functions discussed in the Checking Data section. Similarly, we
provide limited functionality for plotting results, but provide examples
of how to build custom visualizations from output, specifically in R.
For the Shiny application, we provide a base output plot, but the
ability to download your results for custom plotting in programs of the
user's choice. Finally, we provide brief summaries of general
applications of \textbf{hatchR} below, but encourage users to visit
articles hosted on the software webpage that extensively outline primary
functions and applications, especially automating the application of
predicting phenology across multiple variables.

\begin{figure}

{\centering \includegraphics[width=5.68in,height=0.5\textheight]{flowchart}

}

\caption{Full workflow for hatchR. Data processes are demonstrated with beige circles and pakcage functions with white rectangles.}\label{fig:unnamed-chunk-3}
\end{figure}

@@

Effective value models were created by Sparks et al. (2019) to implement
developmental models in wild environments for Sockeye Salmon (\emph{O.
nerka}). The need for their development arose because historic models,
specifically those in Beacham \& Murray (1990), only considered the
average incubation temperature during development and, for wild fishes,
average incubation temperature was impossible to estimate because it was
unknown when fish hatched even if adult spawn timing was known. To
address this, Sparks et al. (2019) used the reciprocal of the
formulation of model 2 from Beacham \& Murray (1990) and assigned an
effective value for every day of development using the daily average
temperature.

@@

Where \emph{i} is the daily value and a fish hatches or emerges when the
cumulative sum of effective values reaches one:
\[\sum_{i =1}^nEffectiveValue_i = 1\]

The effective value model framework is the basis for the phenological
models in \textbf{hatchR}, both in the included \texttt{model\_table} in
the package (though \texttt{model\_table} includes more complex models
developed by Beacham \& Murray (1990)), as well as for custom models
users can fit with \texttt{fit\_model()}. Specifically,
\texttt{model\_table} includes parameterizations from Beacham \& Murray
(1990), Sparks et al. (2017), and Austin et al. (2019) (who extended
McPhail \& Murray (1979)).

@@

Water temperature datasets collected for wild environments are often
either 1.) already summarized by day (\emph{i.e.}, mean daily
temperature) or, 2.) in a raw format from something like a HOBO TidbiT
where readings are taken multiple times per day, which can be summarized
into a mean daily temperatures. Alternatively, new statistical models
like that of Siegel, Fullerton, FitzGerald, Holzer, \& Jordan (2023)
could be similarly implemented.

Fundamentally, \textbf{hatchR} assumes you have input data with two
columns: a date column, giving the date (and often time) of a
temperature measurement, and a temperature column, giving the associated
temperature measurement (in centigrade). Other columns are okay to
include, but these two columns (with any column name---just
\emph{without} spaces) are required. We expect your data to look
something like this:

\begin{longtable}[]{@{}ll@{}}
\toprule\noalign{}
date & temperature \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
2000-01-01 & 2.51 \\
\ldots{} & \ldots{} \\
2000-07-01 & 16.32 \\
\ldots{} & \ldots{} \\
2000-12-31 & 3.13 \\
\end{longtable}

\textbf{hatchR} assumes you've checked for missing records or errors in
your data as it \emph{will function with gaps}, so it's important to go
through the data checks discussed below, as well as your own validity
checks. \textbf{hatchR} can use values down to freezing (e.g, 0 °C),
which returns extremely small effective values, and time to hatch or
emerge may be \textgreater{} 1 year. In these cases, we suggest users
consider how much of that data type is reasonable with their data.

For users choosing to implement \textbf{hatchR} in R, data can be
imported from any format the user chooses, as long as users can
eventually coerce their data into a \texttt{dataframe} or
\texttt{tibble} format, in which each row represents a single record.
For the Shiny application, users must have their data stored as a .csv
(comma separated values) file for upload, which can easily be exported
using datasheet software like Microsoft Excel or Google Sheets.

@@

\textbf{hatchR} is built assuming data will be analyzed as daily average
temperatures. Despite that assumption, raw data (\emph{e.g.}, as
outputted by HOBO loggers) can be used and \textbf{hatchR} includes
functionality to summarize those data into a format that is usable (only
in R, it must be summarized for the Shiny app), as well as provides
functions for basic visual and programmatic data checks to make sure
outliers or missing data are at least brought to user's attention.

We demonstrate the utility of the summary and check functions
\texttt{summarize\_temp()}, \texttt{plot\_check\_temp()}, and
\texttt{check\_continuous()} using a simulated year-long dataset with
temperature readings every thirty minutes.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# create date object for a year with 30 min reading intervals}
\NormalTok{dates }\OtherTok{\textless{}{-}} \FunctionTok{seq}\NormalTok{(}\AttributeTok{from =} \FunctionTok{ymd\_hms}\NormalTok{(}\StringTok{"2000{-}07{-}01 00:00:00"}\NormalTok{),}
             \AttributeTok{to =} \FunctionTok{ymd\_hms}\NormalTok{(}\StringTok{"2001{-}06{-}30 23:59:59"}\NormalTok{), }\AttributeTok{length.out =} \DecValTok{17568}\NormalTok{) }

\CommentTok{\# create empty dataframe}
\NormalTok{year\_sim }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\FunctionTok{matrix}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\AttributeTok{nrow =} \FunctionTok{length}\NormalTok{(dates), }\AttributeTok{ncol =} \DecValTok{1}\NormalTok{)) }

\CommentTok{\# date column}
\FunctionTok{colnames}\NormalTok{(year\_sim)[}\DecValTok{1}\NormalTok{] }\OtherTok{\textless{}{-}} \StringTok{"date"}

\CommentTok{\# add dates vector to date column}
\NormalTok{year\_sim[}\DecValTok{1}\NormalTok{] }\OtherTok{\textless{}{-}}\NormalTok{ dates }

\CommentTok{\#random seed}
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}

\CommentTok{\# take temps from a random normal dist with mean 10 sd 3 }
\CommentTok{\# for every date time combo in dates and append to column (temp) }
\CommentTok{\# in year\_sim}
\NormalTok{year\_sim}\SpecialCharTok{$}\NormalTok{temp }\OtherTok{\textless{}{-}} \FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =} \FunctionTok{length}\NormalTok{(dates), }\AttributeTok{mean =} \DecValTok{10}\NormalTok{, }\AttributeTok{sd =} \DecValTok{3}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{abs}\NormalTok{() }

\FunctionTok{dim}\NormalTok{(year\_sim)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 17568     2
\end{verbatim}

First, we recommend checking imported data for any outliers or strange
inputs using \texttt{plot\_check\_temp()}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{plot\_check\_temp}\NormalTok{(}\AttributeTok{data =}\NormalTok{ year\_sim,}
                \AttributeTok{dates =}\NormalTok{ date,}
                \AttributeTok{temperature =}\NormalTok{ temp,}
                \AttributeTok{temp\_min =} \DecValTok{0}\NormalTok{, }\CommentTok{\# temp\_min and max lines are }
                              \CommentTok{\# user customizable}
                \AttributeTok{temp\_max =} \DecValTok{25}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Full output of the plot\_check\_temp() visual data check
function from hatchR. Note that the red and blue lines are customizable
by the user to create outlier thresholds for their visual check.}
\end{figure}

There are no obvious outliers but since each day has 48 records, we need
to summarize it to daily mean temperature with
\texttt{summarize\_temp()} and then check for missing days with
\texttt{check\_continuous()}. We also recommend using
\texttt{plot\_check\_temp()} again on the summarized data (though leave
out the resulting plot for space efficiency in this manuscript).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# summarize}
\NormalTok{year\_sim\_summ }\OtherTok{\textless{}{-}} \FunctionTok{summarize\_temp}\NormalTok{(}\AttributeTok{data =}\NormalTok{ year\_sim,}
                                \AttributeTok{dates =}\NormalTok{ date,}
                                \AttributeTok{temperature =}\NormalTok{ temp)}

\CommentTok{\# now a year\textquotesingle{}s worth of single{-}day data}
\FunctionTok{dim}\NormalTok{(year\_sim\_summ)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 365   2
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# check continuous (no errors)}
\FunctionTok{check\_continuous}\NormalTok{(}\AttributeTok{data =}\NormalTok{ year\_sim\_summ,}
                 \AttributeTok{dates =}\NormalTok{ date)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## i No breaks were found. All clear!
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# we can demonstrate an error by removing Oct. 8 (100th day)}
\FunctionTok{check\_continuous}\NormalTok{(}\AttributeTok{data =}\NormalTok{ year\_sim\_summ[}\SpecialCharTok{{-}}\DecValTok{100}\NormalTok{,],}
                 \AttributeTok{dates =}\NormalTok{ date)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: ! Data not continuous
## i Breaks found at rows:
## i 100
\end{verbatim}

@@

Users can either use Salmonid models from \texttt{model\_table} included
in the package and the Shiny app. As discussed, these models are
included because parameterizations already existed in the literature.
However, these parameterizations are limited to Pacific Salmon and Bull
Trout. To widen the application of the effective value approach, we
include a \texttt{fit\_model()} function, which is species agnostic (as
long as development generally follows the power law).

The function \texttt{fit\_model()} uses data in which average incubation
temperature (°C) and days to phenological event (as two vectors) are the
inputs. The function estimates parameter coefficients for
\emph{log\textsubscript{e}a} and \emph{b} using \texttt{stats::nls()}

\textbf{BRYAN TO FINISH THIS PARAGRAPH OFF HERE}

\paragraph{Fitting models for other
fishes}\label{fitting-models-for-other-fishes}

Below, we demonstrate how the \texttt{fit\_model()} function may be used
to create custom parameterizations for species beyond the Salmonids in
the \texttt{model\_table} included in the package. We include
parameterizations from three warm-water species to demonstrate the
\texttt{fit\_model()} utility for fishes beyond the scope of the
original effective value approach. These parameterizations are for
commonly cultured sportfishes including Smallmouth Bass
(\emph{Micropterus dolomieu}) from Webster (1948), Channel Catfish
(\emph{Ictalurus punctatus}) from Small \& Bates (2001), and Lake
Sturgeon (\emph{Acipenser fulvescens}) from Smith \& King (2005). We
provide parameterization for just Smallmouth Bass here for concision,
but the full code set for all species is available at
\url{https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd}.

We demonstrate the utility of this approach by creating a random thermal
regime with an ascending thermograph with a mean temperature of 16 °C,
parameterizing models for each species, and demonstrating days to hatch
and developmental period for each species with the random thermal regime
(Figure 3).

\begin{Shaded}
\begin{Highlighting}[]
\DocumentationTok{\#\#\#  make temp regime}
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
\CommentTok{\# create random temps and corresponding dates}
\NormalTok{temps\_sim }\OtherTok{\textless{}{-}} \FunctionTok{sort}\NormalTok{(}\FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =}\DecValTok{30}\NormalTok{, }\AttributeTok{mean =} \DecValTok{16}\NormalTok{, }\AttributeTok{sd =} \DecValTok{1}\NormalTok{), }\AttributeTok{decreasing =} \ConstantTok{FALSE}\NormalTok{)}
\NormalTok{dates\_sim }\OtherTok{\textless{}{-}}  \FunctionTok{seq}\NormalTok{(}\AttributeTok{from =} \FunctionTok{ymd}\NormalTok{(}\StringTok{"2000{-}07{-}01"}\NormalTok{),}
             \AttributeTok{to =} \FunctionTok{ymd}\NormalTok{(}\StringTok{"2000{-}07{-}31"}\NormalTok{), }\AttributeTok{length.out =} \DecValTok{30}\NormalTok{)}

\NormalTok{data\_sim }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\DecValTok{30}\NormalTok{, }\DecValTok{2}\NormalTok{) }\SpecialCharTok{|\textgreater{}} \FunctionTok{data.frame}\NormalTok{()}
\NormalTok{data\_sim[,}\DecValTok{1}\NormalTok{] }\OtherTok{\textless{}{-}}\NormalTok{ temps\_sim}
\NormalTok{data\_sim[,}\DecValTok{2}\NormalTok{] }\OtherTok{\textless{}{-}}\NormalTok{ dates\_sim}

\CommentTok{\# change names so they aren\textquotesingle{}t the same as the vector objects}
\FunctionTok{colnames}\NormalTok{(data\_sim) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"temp\_sim"}\NormalTok{, }\StringTok{"date\_sim"}\NormalTok{)}

\DocumentationTok{\#\#\# smallmouth mod}
\NormalTok{smallmouth }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{2}\NormalTok{) }\SpecialCharTok{|\textgreater{}} \FunctionTok{data.frame}\NormalTok{()}
\FunctionTok{colnames}\NormalTok{(smallmouth) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"hours"}\NormalTok{, }\StringTok{"temp\_F"}\NormalTok{)}
\NormalTok{smallmouth}\SpecialCharTok{$}\NormalTok{hours }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{52}\NormalTok{, }\DecValTok{54}\NormalTok{, }\DecValTok{70}\NormalTok{, }\DecValTok{78}\NormalTok{, }\DecValTok{90}\NormalTok{, }\DecValTok{98}\NormalTok{, }\DecValTok{150}\NormalTok{, }\DecValTok{167}\NormalTok{, }\DecValTok{238}\NormalTok{, }\DecValTok{234}\NormalTok{)}
\NormalTok{smallmouth}\SpecialCharTok{$}\NormalTok{temp\_F }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{77}\NormalTok{, }\DecValTok{75}\NormalTok{, }\DecValTok{71}\NormalTok{, }\DecValTok{70}\NormalTok{, }\DecValTok{67}\NormalTok{, }\DecValTok{65}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{59}\NormalTok{, }\DecValTok{55}\NormalTok{, }\DecValTok{55}\NormalTok{)}

\CommentTok{\# change F to C and hours to days}
\NormalTok{smallmouth }\OtherTok{\textless{}{-}}\NormalTok{ smallmouth }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{days =} \FunctionTok{ceiling}\NormalTok{(hours}\SpecialCharTok{/}\DecValTok{24}\NormalTok{),}
         \AttributeTok{temp\_C =}\NormalTok{ (temp\_F }\SpecialCharTok{{-}}\DecValTok{32}\NormalTok{) }\SpecialCharTok{*}\NormalTok{ (}\DecValTok{5}\SpecialCharTok{/}\DecValTok{9}\NormalTok{))}

\NormalTok{smb\_mod }\OtherTok{\textless{}{-}} \FunctionTok{fit\_model}\NormalTok{(}\AttributeTok{temp =}\NormalTok{ smallmouth}\SpecialCharTok{$}\NormalTok{temp\_C,}
                     \AttributeTok{days =}\NormalTok{ smallmouth}\SpecialCharTok{$}\NormalTok{days,}
                     \AttributeTok{species =} \StringTok{"smb"}\NormalTok{,}
                     \AttributeTok{development\_type =} \StringTok{"hatch"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Note the \emph{R\textsuperscript{2}} fit from the models below. You can
see they generally all preform well and are in line with values from
model 2 of Beacham \& Murray (1990).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#model fits}
\NormalTok{smb\_mod}\SpecialCharTok{$}\NormalTok{r\_squared; cat\_mod}\SpecialCharTok{$}\NormalTok{r\_squared; sturgeon\_mod}\SpecialCharTok{$}\NormalTok{r\_squared}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 0.9868067
\end{verbatim}

\begin{verbatim}
## [1] 0.9433598
\end{verbatim}

\begin{verbatim}
## [1] 0.9217358
\end{verbatim}

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/unnamed-chunk-10-1.pdf}
\caption{Days to hatch as predicted using custom parameterizations for
Channel Catfish (blue), Lake Sturgeon (black, and Smallmouth Bass
(green) as generated by the fit\_model() function in hathcR. Models are
all run using a spawn date starting June 1 for a randomly generated
ascending thermal regime with mean temperature 16 °C}
\end{figure}

@@

To illustrate model selection and phenology prediction we will recreate
a small portion of the analysis done by Sparks et al. (2019) using the
\texttt{woody\_island} dataset included in this package . We will
predict both hatch and emergence timing, so we will obtain a model
expression for each using \texttt{model\_select()}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sockeye\_hatch\_mod }\OtherTok{\textless{}{-}} \FunctionTok{model\_select}\NormalTok{(}
  \AttributeTok{author =} \StringTok{"Beacham and Murray 1990"}\NormalTok{, }
  \AttributeTok{species =} \StringTok{"sockeye"}\NormalTok{, }
  \AttributeTok{model =} \DecValTok{2}\NormalTok{, }
  \AttributeTok{development\_type =} \StringTok{"hatch"}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

We can now use our model expressions to predict when sockeye would hatch
and emerge at Woody Island in 1990. First we predict hatch timing using
\texttt{predict\_phenology()}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{WI\_hatch }\OtherTok{\textless{}{-}} \FunctionTok{predict\_phenology}\NormalTok{(}
  \AttributeTok{data =}\NormalTok{ woody\_island,}
  \AttributeTok{dates =}\NormalTok{ date,}
  \AttributeTok{temperature =}\NormalTok{ temp\_c,}
  \AttributeTok{spawn.date =} \StringTok{"1990{-}08{-}18"}\NormalTok{,}
  \AttributeTok{model =}\NormalTok{ sockeye\_hatch\_mod}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

And then look inside the returned object to see days to hatch and
development period:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{WI\_hatch}\SpecialCharTok{$}\NormalTok{days2done}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## NULL
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{WI\_hatch}\SpecialCharTok{$}\NormalTok{dev.period}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##        start       stop
## 1 1990-08-18 1990-10-30
\end{verbatim}

\paragraph{Understanding your results}\label{understanding-your-results}

The output from \texttt{predict\_phenology()} includes a lot of
information. If we look at our \texttt{WI\_hatch} object we see there
are multiple elements stored in a list which can be accessed using the
\texttt{\$} operator.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{str}\NormalTok{(WI\_hatch)}
\end{Highlighting}
\end{Shaded}

\texttt{WI\_hatch\$days2done} outputs the predicted days to develop.
